---
title: "Class 09: Candy Mini-Project"
author: "Erin Jang (PID: A17713992)"
format: pdf
toc: true
---

## Background

In this mini-project, you will explore FiveThirtyEight’s Halloween Candy dataset. 

Today we are taking a detor to analyze a fun dataset (that we have more intrinsic insight into) with the most useful analysis method we have learned thus far - Principal Component Analysis (PCA). 

## Importing candy data 
The data is all related to halloween candy and is from the 538 website. 

```{r}
candy_file <- "candy-data.txt"

candy = read.csv(candy_file, row.names = 1)
head(candy)
```

### What is in the dataset?

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```
There are 85 different candy types. 

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```
There are 38 fruity candy types. 

### What is your favorite candy?

```{r}
candy["Twix", ]$winpercent
```

> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
candy["3 Musketeers", ]$winpercent
```
My favorite candy is 3 Musketeers and its winpercent value is 67.60294. 

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```
The winpercent value for Kit Kat is 76.7686.

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```
The winpercent value for Tootsie Roll Snack Bars is 49.6535. 

There is a useful `skim()` package. 

```{r}
# install.packages("skimr")
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The sugarpercent looks to be on a different scale as it represents the percentile of the sugar content, not a value of 0 or 1. 

> Q7. What do you think a zero and one represent for the candy$chocolate column?

A zero represents that the candy is not a chocolate, whereas a one represents that the candy is a chocolate. 

## Exploratory analysis

> Q8. Plot a histogram of winpercent values using both base R and ggplot2.

Histogram using base R:
```{r}
hist(candy$winpercent)
```

Histogram using ggplot2: 
```{r}
library(ggplot2)
ggplot(candy, aes(winpercent)) +
  geom_histogram(bins = 20)
```

> Q9. Is the distribution of winpercent values symmetrical?

No, the distribution of winpercent values is not symmetrical. 

> Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```
If we calculate by mean, the center of the distribution is slightly above 50%. However, if we calculate by median, which seems to be more appropriate based on the histogram, the center of the distribution is slightly below 50%. 

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])
mean(candy$winpercent[as.logical(candy$fruit)])

## choc.ind <- as.logical(candy$chocolate)
## choc.candy <- candy[choc.ind, ]
## choc.win <- choc.candy$winpercent
## mean(choc.win)
```
On average, chocolate candy is higher ranked than fruit candy. 

> Q12. Is this difference statistically significant?

```{r}
t.test(
  candy$winpercent[as.logical(candy$chocolate)],
  candy$winpercent[as.logical(candy$fruit)]
)
```
This difference is statistically significant as the p-value is less than 0.05. 

## Overall candy rankings

> Q13. What are the five least liked candy types in this set?

```{r}
head(candy[ order(candy$winpercent), ], 5)
```
The five least liked candies are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters. 

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
tail(candy[ order(candy$winpercent), ], 5)
```
The top 5 all time favorite candy types out of this set are Reese's Peanut Butter cup, Reese's Miniatures, Twix, Kit Kat, and Snickers. 

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r, fig.height = 15, fig.width = 10}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r, fig.height = 15, fig.width = 10}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```

### Time to add some useful color

```{r, fig.height = 15, fig.width = 10}
my_cols = rep("black", nrow(candy))
## chocolate candy in chocolate color
my_cols[as.logical(candy$chocolate)] = "chocolate"
## candy bars in brown
my_cols[as.logical(candy$bar)] = "brown"
## fruity candy in pink
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) +
  ylab("")
```

> Q17. What is the worst ranked chocolate candy?

Sixlets is the worst ranked chocolate candy. 

> Q18. What is the best ranked fruity candy?

Starbursts is the best ranked fruity candy. 

## Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of win vs price
ggplot(candy) +
  aes(x = winpercent, y = pricepercent, label = rownames(candy)) +
  geom_point(col = my_cols) + 
  geom_text_repel(col = my_cols, size = 3.3, max.overlaps = 25)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
head(candy[order(-candy$winpercent / candy$pricepercent), ], 5)
```
Tootsie Roll Midgies is the highest ranked in terms of winpercent for the least money. 

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
top5 <- head( candy[ord,c(11,12)], n=5 )

top5[order(top5$winpercent), ]
```
The least popular of the top 5 most expensive candy types in the dataset are Nik L Nip, Ring pop, Nestle Smarties, Hershey's Milk Chocolate, and Hershey's Krackel. 

### Optional

```{r, fig.height = 15, fig.width = 10}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

## Exploring the correlation structure

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity are strongly anti-correlated. This makes sense because candies are typically one or the other, not both.

> Q23. Use your corrplot result to make a prediction: which variables do you expect will have the largest contributions (i.e. loadings) to PC1 (i.e., drive the most separation between candies along the first principal component)?

I expect chocolate, bar, and peanutyalmondy to have strong positive loadings, and fruity, hard, and pluribus to have strong negative loadings on PC1, showing a major separation between chocolate candies and fruity candies.

## Principal Component Analysis

In this case we want to be sure to set `scale = TRUE` argument for `prcomp()` because we have one variable, `winpercent`, that is on a very difference scale than all others and would otherwise dominate our PCA results. 

```{r}
candy_file <- "candy-data.txt"
candy = read.csv(candy_file, row.names = 1)
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

First major result figures is the "score plot" of PC1 vs PC2 - how he different candy are related to each other on our new PC axis:
```{r}
plot(pca$x[,1:2])
```
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```
```{r}
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x = PC1, y = PC2, 
            size = winpercent/100,  
            text = rownames(my_data),
            label = rownames(my_data)) +
        geom_point(col = my_cols)
p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 40)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```
```{r}
## library(plotly)
## ggplotly(p)
```

The second major results figure from PCA is the so-called "loadings plot".

> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

```{r}
ggplot(pca$rotation) +
  aes(x = PC1,
      reorder(row.names(pca$rotation), PC1)) +
  geom_col()
```
The variables fruity, pluribus, and hard were picked up strongly by PC1 in the positive direction. This makes sense as this relationship was previously highlighted when looking at the correlation structure graph.  

## Summary

> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

A “winning” candy in this dataset is typically chocolate, especially in bar form, frequently with ingredients such as peanuts, caramel, or nougat. These candies tend to have higher winpercent values, meaning they are more consistently preferred. The exploratory analysis supports this by showing that chocolate candies have higher average winpercent than fruit ones. In addition, the correlation analysis further emphasizes that chocolate is positively associated with winpercent as well as bar and peanutyalmondy, while fruity has a negative correlation. Finally, the PCA results shows that the PC1 separates candies along a chocolate vs fruity axis. Variables including chocolate, bar, and peanutyalmondy move strongly in the positive direction, while fruity and hard go in the opposite direction. 