---
title: "Class 08: Breast Cancer Mini Project"
author: "Erin Jang (PID: A17713992)"
format: pdf
toc: true
---

## Background

The goal of this mini-project is for you to explore a complete analysis using the unsupervised learning techniques covered in class.

Today we will analyze a data-set from fine needle aspiration (FNA) of a breast mass.

## Data Import

The data is made available as a CSV file for download. We can read this in using `read.csv()`:

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names = 1)
```

```{r}
head(wisc.df, 3)
```

Make sure we remove or exclude the `diagnosis` column from the data-set that we use for further analysis - this is the expert diagnosis as either M or B. 

```{r}
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]
diagnosis <- as.factor( wisc.df$diagnosis ) 
```

## Exploratory data analysis

> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.data)
```
There are 569 observations. 

> Q2. How many of the observations have a malignant diagnosis?

```{r}
table(wisc.df$diagnosis)
```
212 observations have a malignant diagnosis. 

> Q3. How many variables/features in the data are suffixed with _mean?

We can use the `grep()` function to help us here:
```{r}
colnames(wisc.data)
```

```{r}
length(grep("_mean", colnames(wisc.data), value = TRUE))
```
There are 10 variables in the data suffixed with _mean. 

## Principal Component Analysis (PCA)

We need to scale our data before PCA with the `scale-TRUE` argument to `prcomp()`.

```{r}
wisc.pr <- prcomp(wisc.data, scale = TRUE)
summary(wisc.pr)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

44.27% is captured by the first principal component. 

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

The first three principal components are required to describe at least 70% of the original variance in the data.

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

The first seven principal components are required to describe at least 90% of the original variance in the data. 

```{r}
biplot(wisc.pr)
```
> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

This plot is incredibly difficult to understand as conclusions cannot be made from the plot with everything crowded together.

```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC3, col=diagnosis) +
  geom_point()
```

For both plots, the malignant observations are towards the left of the graph whereas the benign are assembled towards the right. This presents a separation between the benign and malignant samples. 

### Variance explained

```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(c(1,pve), xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

```{r}
## ggplot based graph
#install.packages("factoextra")
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

### Communicating PCA results

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

```{r}
wisc.pr$rotation["concave.points_mean", 1]
```

The loading of concave.points_mean on PC1 is −0.2609, which indicates a strong contribution to the first principal component. There are other features with larger contributions, such as radius_mean. 

## Hierarchical clustering

```{r}
# Scale the wisc.data data using the "scale()" function
# Next, calculate the (Euclidean) distances between all pairs of observations in the new scaled dataset and assign the result to data.dist.
wisc.hclust <- hclust(dist(scale(wisc.data)))
plot(wisc.hclust)
```

> Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h = 19, col = "red", lty = 2)
```

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)
table(wisc.hclust.clusters, diagnosis)
```

> Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

The ward.D2 method gives the best results for this dataset. It produces more clearly separated clusters compared to methods like "single", "complete", and "average". Because ward.D2 minimizes variance within clusters, it forms spherical clusters, which aligns well with the structure of this data and results in better separation of malignant and benign samples.

## Combining methods

```{r}
d <- dist(wisc.pr$x[,1:7])
wisc.pr.hclust <- hclust(d, method = "ward.D2")
plot(wisc.pr.hclust)
abline(h = 70, col = "red")
```

```{r}
grps <- cutree(wisc.pr.hclust, k = 2)
table(grps)
```

How does this clustering `grps` correspond to the expert `diagnosis`?

```{r}
table(grps, diagnosis)
```

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC2) +
  geom_point(col=grps)
```

```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
```

> Q13. How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?

```{r}
table(wisc.pr.hclust.clusters, diagnosis)
```
The newly created hclust model with two clusters separate out the two diagnoses pretty well with only 52 miscalculations. This shows an approximately 91% accuracy. 

> Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses.

```{r}
table(wisc.hclust.clusters, diagnosis)
```
The hierarchical clustering model without PCA performs worse at separating malignant and benign samples. The clusters are more fragmented and show greater mixing of diagnoses.

## Prediction

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```
```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

> Q16. Which of these new patients should we prioritize for follow up based on your results?

Patient 1 should be prioritized for follow up based on these results, as their PCA projection is closer to the malignant cluster, while Patient 2 appears more consistent with benign samples.

## About this document

```{r}
sessionInfo()
```

