---
title: "Class 12: Introduction to Genome Informatics"
author: "Erin Jang (PID: A17713992)"
format: pdf
toc: true
---

# Section 1. Proportion og G/G in a population

Downloaded a CSV file from Ensemble <https://www.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39895045-39895145;v=rs8067378;vdb=variation;vf=959672880;sample=MXL#373531_tablePanel>

Here we read this CSV file:

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

## Section 4: Population Scale Analysis (Homework)

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
table(expr$geno)
```

```{r}
median(expr$exp[expr$geno == "A/A"])
median(expr$exp[expr$geno == "A/G"])
median(expr$exp[expr$geno == "G/G"])
```

The sample size for A/A genotype is 108, 233 for A/G genotype, and 121 for G/G genotype. Their median expression levels for the A/A, A/G, and G/G genotypes are 31.25, 25.06, and 20.07, respectively. 

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

Let's make a boxplot:

```{r}
library(ggplot2)
```

```{r}
boxplot <- ggplot(expr) + aes(geno, exp, fill = geno) +
  geom_boxplot(notch = TRUE)
boxplot
```

The boxplot shows that individuals with the A/A genotype have much higher ORMDL3 expression levels compared to individuals with the G/G genotype. This suggests that the SNP is associated with differences in ORMDL3 expression, supporting that the SNP affects ORMDL3 gene expression.
