---
title: "Class 05: Data Visualization with ggplot2"
author: "Erin (PID: A17713992)"
format: pdf
toc: TRUE
---

## Background

There are many graphics systems in R for making plots and figures. These include so-called *"base R" graphics* like the `plot()` function and add on packages like **ggplot2**.

Let's compare how we make a simple figure with these two systems. 

We can use the in-built `cars` dataset:

```{r}
head(cars)
```

```{r}
plot(cars)
```

Before I can use ggplot 2 I need to install it on my computer.
To do this we can use the function `install.packages("ggplot2")`

>**N.B.** We never run `install.packages()` in our quarto doc (we run it only once in our R console) as it would re-install the package every time we render our quarto report.

Once installed we need to load up the package into our R brain:
```{r}
library(ggplot2)
```

The main function in the **ggplot2** package is called `ggplot()` 

```{r}
ggplot(cars)
```

Every ggplot has at least 3 layers:

- the **data** (a data.frame of the stuff we want to plot)
- the **aes**thetics (how the data maps to the plot)
- the **geom** layer (how you want the plot drawn, e.g. points, lines, etc.)

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point()
```

## Add some custom features

Let's add a trend line that shows the relationship between speed and distance. 

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point() +
  geom_line()
```

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth() + 
  theme_bw() + 
  labs(title="Stopping distance of old cars",
       x="Speed (MPH)",
       y="Distance (ft)")
```


Q. Can you make the `geom_smooth()` function produce a linear straight line fit to the data and trn off the "gray" error region?

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method="lm", se = FALSE) + 
  theme_bw() + 
  labs(title="Stopping distance of old cars",
       x="Speed (MPH)",
       y="Distance (ft)")
```


---

## Gene Expression Figure

Import the data to plot

```{r}  
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
sum(genes$State == "up")
```

A useful new function in this context is the `table()` function:

```{r}
table(genes$State)
```

My first plot attempt

```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col=State) +
  geom_point()
```

```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col=State) +
  geom_point() +
  labs(title = "Expression changes upon GLP-1 inhibitor treatment ", x = "No drug", y = "Drug") +
  scale_colour_manual(values=c("blue","gray","red")) + 
  theme_bw()
```

## Going further

Here we read the famous gapminder dataset:

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

> Q. How many entires (i.e. rows) are in this dataset?

```{r}
nrow(gapminder)
```

> Q. How many different "country" entries are in this dataset?

```{r}
length(unique(gapminder$country))
```

Let's make our first plot of the entire dataset:

Plot of "gdpPercap" vs "lifeExp" colored by "continent"

```{r}
p <- ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col= continent) + 
  geom_point(alpha = 0.3)
``` 

```{r}
p
```

I can add more layers to `p`
```{r}
p + 
  facet_wrap(~gapminder$continent)
```

Make a plot for 1977 and 2007 only (not all the years in the dataset).

> Q. First use the **dplyr** package and the `filter()` function from that package to extract the rows from the year 2007. 

```{r, message=FALSE, warning=FALSE}
library(dplyr)
```

```{r}
g07 <- filter(gapminder, year == 2007)
g77 <- filter(gapminder, year == 1977)
g <- filter(gapminder, year == 2007 | year == 1977)
```

```{r}
ggplot(g) +
  aes(gdpPercap, lifeExp, col = continent, size = pop) + 
  geom_point() +
  facet_wrap(~year)
```

> Q. Make a histogram of lifeExp colored by continent (try using `fill=continent` or `col=continent`)

```{r}
ggplot(gapminder) +
  aes(lifeExp, fill=continent) +
  geom_histogram()
```

> Q. Make a histogram of lifeExp faceted by continent

```{r}
ggplot(gapminder) +
  aes(lifeExp) +
  geom_histogram() +
  facet_wrap(~continent)
```

